---
title: "The scale of a substitution rate matrix"
output: html_notebook
---

*In this notebook, we will try to see how the scale of a substitution rate matrix matters.*

```{r}
library(expm) # to compute a matrix exponential
```

## Useful functions for simulating DNA evolution along a branch
```{r}
# Function for drawing a DNA state, for instance at the root
# We assume the order is A, C, G, T
drawStateAtRoot <- function (probabilities){
    states <- c("A", "C", "G", "T")
    u <- runif(n=1, min=0,max=1)
    if (u < probabilities[1])
      { return (states[1])}
    else if (u < probabilities[1]+probabilities[2])
      { return (states[2])}
    else if (u < probabilities[1]+probabilities[2]+probabilities[3])
      { return (states[3])}
    else { return (states[4])}
}

# Function to draw a state on a branch
drawNewState <- function (probabilities, states){
    u <- runif(n=1, min=0,max=1)
    if (u < probabilities[1])
      { return (states[1])}
    else if (u < probabilities[1]+probabilities[2])
      { return (states[2])}
    else { return (states[3])}
}

# Function to compute the stationary frequencies associated to a rate matrix
computeStationaryFrequencies <- function(rateMatrix) {
  return(expm(rateMatrix*100)[1,])
}

# Function to simulate along a branch using the Gillespie algorithm,
# under a particular rate matrix, given some branch length, 
# and given a starting state.
simulateAlongBranch <- function(startingState, branchLength, rateMatrix) {
    l <- 0.0
    numberOfSubstitutions <- 0
    current <- startingState
    while (l < branchLength) {
        currentRow=1
        if (current =='C') {
          currentRow=2  
        }
        else if (current =='G') {
          currentRow=3  
        }
        else if (current =='T') {
          currentRow=4  
        }
        # rateOfMovingAway is the sum of the rates of all possible events, 
        # aka minus the diagonal element
        rateOfMovingAway = - rateMatrix[currentRow, currentRow]
        l <- l + rexp(rate=rateOfMovingAway, n=1) 
        if (l < branchLength) {
            numberOfSubstitutions <- numberOfSubstitutions + 1
            possibleArrivingIndices <- setdiff(c(1,2,3,4), currentRow)
            probabilitiesOfArrivingStates <- rateMatrix[currentRow, possibleArrivingIndices]/rateOfMovingAway
            current <- drawNewState(probabilitiesOfArrivingStates, c("A", "C", "G", "T")[possibleArrivingIndices])
        }
    }
    return (c(current, numberOfSubstitutions))
}

# Simulate a site history given starting (root) frequencies, a branch length,
# and a rate matrix.
simulateSiteHistory <- function (startingFrequencies, branchLength, rateMatrix) {
  rootState <- drawStateAtRoot(startingFrequencies)
  return(simulateAlongBranch(startingState=rootState, branchLength=branchLength, rateMatrix=rateMatrix))
}

# Simulate numberOfSites site histories along a branch of length branchLength,
# with a rate matrix rateMatrix.
simulateManySiteHistories <- function (branchLength, rateMatrix, numberOfSites) {
  stationaryFrequencies <- computeStationaryFrequencies(rateMatrix)
  allHistories <- matrix(rep(NA, 2*numberOfSites), nrow=numberOfSites, ncol=2)
  for (i in 1:numberOfSites) {
    allHistories[i,] <- simulateSiteHistory(startingFrequencies=stationaryFrequencies, branchLength=branchLength, rateMatrix=rateMatrix)
  }
  return(allHistories)
}

# Plot the number of substitutions as contained in siteHistories.
plotNumbersOfSubstitutions <- function (siteHistories) {
  numbers <- as.numeric(siteHistories[,2])
  hist(numbers, main=paste0("Number of substitutions; mean = ", mean(numbers)), xlab="Numbers of substitutions", ylab="Number of sites")
}

```


## Let's create a substitution rate matrix, and look at it
```{r}
substitutionRateMatrix <- 3*matrix(c(c(-1.916, 0.541, 0.787, 0.588), c(0.148, -1.069, 0.415, 0.506), c(0.286, 0.170, -0.591, 0.135), c(0.525, 0.236, 0.594, -1.355)), nrow=4, ncol=4, byrow=TRUE)
print(substitutionRateMatrix)
```

## Let's simulate 1000 sites with this substitution matrix
We want to simulate DNA evolution along a branch of length 0.5.
```{r}
siteHistories <- simulateManySiteHistories(branchLength=0.5, rateMatrix=substitutionRateMatrix, numberOfSites=1000)
```

### While doing our simulation, we counted the number of substitutions per site
Let's look at them!
```{r}
plotNumbersOfSubstitutions(siteHistories)
```


When simulating along a branch of length 0.5 with our substitution rate matrix, on average there are ~1.5 substitution per site. That's not very convenient.


## Scaling the substitution rate matrix
Let's scale the substitution rate matrix: at stationarity, we want the total rate of substitution per unit time to be 1, i.e.:

$$\sum_{i \in {A,C,G,T}}~\sum_{j \in {A,C,G,T}} \pi_i q_{ij} = 1,~~i \neq j$$
or, equivalently:
$$-\sum_{i \in {A,C,G,T}} \pi_i q_{ii} = 1$$

What is the scale of our current matrix?

```{r}
computeScaleOfMatrix <- function(rateMatrix) {
  stationaryFrequencies <- computeStationaryFrequencies(rateMatrix)
  qiis <- c(rateMatrix[1,1], rateMatrix[2,2], rateMatrix[3,3], rateMatrix[4,4])
  scale <- -(stationaryFrequencies[1] * qiis[1] + stationaryFrequencies[2] * qiis[2] + stationaryFrequencies[3] * qiis[3] + stationaryFrequencies[4] * qiis[4])
  return(scale)
}
```

```{r}
scale <- computeScaleOfMatrix(substitutionRateMatrix)
print(scale)
```

Our matrix does not have scale 1.0...

## Rescaling our matrix

```{r}
scaledSubstitutionRateMatrix <- substitutionRateMatrix / scale
```

## Simulating with a scaled substitution matrix:

```{r}

siteHistories <- simulateManySiteHistories(branchLength=0.5, rateMatrix=scaledSubstitutionRateMatrix, numberOfSites=1000)
plotNumbersOfSubstitutions(siteHistories)

```
When we simulate sites along a branch of length 0.5, with a scaled substitution rate matrix, we get on average 0.5 substitution per site.


# Conclusion
To be able to interpret branch lengths as expected numbers of substitutions per site, we need to use scaled substitution rate matrices.



